---
title: "Reproducible research project 1"
author: "Yue Dai"
date: "`r Sys.Date()`"
output:
  prettydoc::html_pretty:
    theme: architect
    highlight: github
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,  warning=FALSE, message=FALSE, fig.width=10, fig.height=5)
```

## Introduction

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as NA)
* **date**: The date on which the measurement was taken in YYYY-MM-DD format
* **interval**: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data

Loading the data

```{r, echo=TRUE}
library(ggplot2)
library(plyr)
activity <- read.csv("~/course project/activity.csv")
str(activity)
```

```{r, echo=TRUE}
activity$date <- as.Date(activity$date) ## transfer factor to date
```

## What is mean total number of steps taken per day?

Calsulate the total number of steps taken per day

```{r, echo=TRUE}
activityclean <- activity[!is.na(activity$steps),] ## remove NAs

## 1. Calculate the total number os steps tanken per day
stepsumd <- aggregate(activityclean$steps ~ activityclean$date, FUN=sum)
colnames(stepsumd)<- c("date", "steps")
```

```{r, echo=TRUE}
## 2. Make a histogram of the total number of steps taken each day
hist(stepsumd$steps, breaks=10, xlab="Steps", main = "Total Steps per Day")
```

```{r, echo=TRUE}
## 3. Calculate and report the mean and median of the total number of steps taken per day
MeanSteps <- mean(stepsumd$steps, na.rm = TRUE)
MeanSteps
```

```{r, echo=TRUE}
MedianSteps <- median(stepsumd$steps, na.rm = TRUE)
MedianSteps
```

## What is the average daily activity pattern?

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r, echo=TRUE}
## 1. Calculating average steps
averageSteps <- aggregate(x=list(avg_steps=activity$steps), by=list(interval=activity$interval), FUN=mean, na.rm=TRUE)
```

```{r, echo=TRUE}
## 2. Plotting
g <- ggplot(averageSteps, aes(x = interval, y = avg_steps))
g + geom_line() + ggtitle("Average Number of Steps per Interval") + 
    xlab("Interval")+ylab("Average Number of Steps")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}
## Maximum steps 
averageSteps[which.max(averageSteps$avg_steps),]
```

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset

```{r, echo=TRUE}
sum(is.na(activity$steps))
```

2. Devise a strategy for filling in all of the missing values in the dataset

We create a new data frame **activity2** that is equal to the original dataset but with the missing data filled in through using mean for that interval for imputation

3. Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r, echo=TRUE}
activity2 <- activity
na <- is.na(activity2$steps)
avg_interval<- tapply(activity2$steps, activity2$interval, mean, na.rm=TRUE, simplify = TRUE)
activity2$steps[na] <- avg_interval[as.character(activity2$interval[na])]
head(activity2)
```

```{r, echo=TRUE}
sum(is.na(activity2$steps))
```

4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day

```{r, echo=TRUE}
## Calculate the total number os steps tanken per day
stepsumd2 <- aggregate(activity2$steps ~ activity2$date, FUN=sum)
colnames(stepsumd2)<- c("date", "steps")
```

```{r, echo=TRUE}
## Make a histogram of the total number of steps taken each day
hist(stepsumd2$steps, breaks=10, xlab="Steps", main = "Total Steps per Day with Filling NAs", col = "Red")
hist(stepsumd$steps, breaks=10, xlab="Steps", main = "Total Steps per Day with Filling NAs", col = "Green", add = T)
legend("topright", c("Imputed Data", "No NA Data"), fill = c("Red", "Green"))
```

```{r, echo=TRUE}
## Calculate and report the mean and median of the total number of steps taken per day
MeanSteps2 <- mean(stepsumd2$steps, na.rm = TRUE)
MeanSteps2
```

```{r, echo=TRUE}
MedianSteps2 <- median(stepsumd2$steps, na.rm = TRUE)
MedianSteps2
```

The new mean is 10766 and the new median is 10766 . Compare with the no NAs data set (mean 10766 and median 10765) , the mean doesn't change, and the median has a small change. It may because the filling strategy we used (filling with means for intervals). From the histogram, we can see now higher frequency counts in the histogram at the center region (close to the mean).

## Are there differences in activity patterns between weekdays and weekends?

```{r, echo=TRUE}
## 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend”
activity2$day <- weekdays(activity2$date)
activity2$datetype <- ifelse(activity2$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
head(activity2)
```

```{r, echo=TRUE}
## 1. Create a new factor variable in the dataset with two levels – “weekday” and “weekend”
activity2$day <- weekdays(activity2$date)
activity2$datetype <- ifelse(activity2$day %in% c("Saturday", "Sunday"), "Weekend", "Weekday")
head(activity2)
```

```{r, echo=TRUE}
## 2. Calculating average steps
averageSteps2 <- aggregate(steps ~ interval + datetype, data = activity2, mean)
```

```{r, echo=TRUE}
## 3. Plotting
t <- ggplot(averageSteps2, aes(x = interval, y = steps, color = datetype))
t + geom_line() + facet_grid(datetype ~ .) +
    ggtitle("Average Number of Steps per Interval on Type of Day") + 
    xlab("Interval")+ylab("Average Number of Steps")
```

From the panel plot we can see the weekday activities start earlier than the weekends. But the weekends have higher activity levels than the weekdays. It may because people have more flexible time during weekend to do activities.